Vigilantism and Institutions:\\ Understanding Attitudes toward Lynching in Brazil

### 2022-11-29 article/article.tex
add paragraph to conclusion

% Packages I've been using for differen reasons
\usepackage{hyperref}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usepackage{mathtools}
\usepackage{caption}

@@

\section{Introduction}
\label{sec:introduction}

@@

The regulation of violence is a crucial function of the state. When states
control violence effectively, markets, politics, and civil society can flourish
\citep{besley2011pillars, north2009violence}. However, states still struggle to
maintain the monopoly of force in much of the Global South. Scholars have
written extensively about how rebel organizations build support to challenge
government institutions, yet we know little about what motivates local
extrajudicial violence and vigilantism in the developing world
\citep{bateson2020politics}.\footnote{On vigilantism more generally, see
\citet{schuberth2013challenging,
zizumbo2017community,smith2019contradictions,cohen2022collective}.} This
omission is significant not only because violence is pervasive in weak states,
but it also comes at a great cost \citep{blair2017predicting}.

One of the most serious forms of extrajudicial violence is lynching, which can
be defined as ``incidents of physical violence committed by large numbers of
private citizens against one or more individuals accused of having committed a
``criminal'' offense, whether or not this violence resulted in the death of the
victim(s)'' \citep[645]{godoy2004justice}. Although lynchings occur in more
than one hundred countries in all regions of the world
\citep{jung2020lynching,smith2019contradictions}, Latin America has been
particularly affected by a sharp increase in vigilante violence. Lynching
episodes have been reported in Guatemala, Mexico, Peru, Venezuela and other
countries in the region \citep{cruz2019determinants, godoy2004justice,
barbara2015vigilantes}. From 2011 to 2015, Brazil registered about 2,500
lynching episodes, and 173 people were killed by angry mobs in 2015
alone---nearly one execution every two days \citep{barbara2015vigilantes,
oliveira2016mob}. According to José de Souza Martins
\citeyearpar{martins2015linchamentos}, who has studied lynchings in Brazil for
more than thirty years, these figures are not only the highest in the country's
history, but among the highest in the world. The people who participate in
lynchings are typically young men, but they also sometimes include teenage
women and girls, elderly women, and even members of the local police
\citep{moura2017linchamentos}.

@@

Our empirical approach offers several distinct advantages. First, the data we
collected come from the same level of analysis as the phenomenon itself: local
residents. Second, online surveys are less susceptible to social desirability
bias, and survey experiments help to elicit sensitive information---an
especially important benefit in researching this topic
\citep{horiuchi2022does}. Third, the experimental nature of the approach allows
us to precisely estimate the size of effects and compare different theoretical
explanations simultaneously. Finally, we can also control for confounding,
which is a potential concern in previous research on vigilantism.

@@

\section{When Is Lynching Perceived as More Justified?}
\label{sec:exp01}

@@

Figure \ref{fig:exp01} shows the main results for the conjoint experiment. The
graph illustrates the preference associated with each attribute of the
hypothetical lynching episode. Dots with horizontal bars represent point
estimates and 95 percent confidence intervals from linear regressions with
robust standard errors clustered at the respondent level.\footnote{Based on a
semisupervized Dirichlet allocation model, we estimated about 20 percent of
respondents were opposed to lynchings in general but forced to choose. Please
refer to Section C.5 of the supplementary material for further information.}

@@

%%%%%%%%%%%%%%%%%%%%%
\section{How to Reduce Support for Lynchings}
\label{sec:exp03}

Even if people view lynchings as a legitimate activity, that does not mean that
this norm cannot be changed. In our second experiment, we measured the effect
of information provision on attitudes about lynching. More specifically, we
tested whether reminding respondents about the legal and social consequences of
vigilante justice reduces subjects' support for it. The experiment included
three treatment conditions and a control group. In all of them we presented
respondents with a short statement affirming that some Brazilians support
vigilantism under certain conditions. We asked the control group to rate their
agreement with the statement. Respondents were asked to use 0 to 49 if they
disagreed, 50 if they neither agreed nor disagreed, and 50--100 if they agreed
with the statement.

@@

\section{Conclusion}
\label{conclusion}

@@

%%%%DELETE OR REVISE In this respect, interventions that reduce popular support for extrajudicial violence can have a substantial impact on citizen welfare and potentially increase state legitimacy in young democracies.

@@

Our work also suggests new avenues for further research. While previous studies
indicate that there is a relationship between pro-lynching attitudes and actual
engagement in vigilante violence \citep{weisburd1988vigilantism}, future
research should try to identify which factors motivate citizens to participate
in lynchings episodes \citep{nivette2016institutional}. Moreover, it is still
unclear whether other forms of vigilantism, such as security patrols, property
damages, or coercive interrogations \citep{bateson2020politics} have similar
levels of popular support. Finally, lynchings have often been described as a
response to low state capacity, yet in many cases state agents actively incite
or engage in vigilante violence themselves \citep{arias2010violent}. Future
studies may investigate in which circumstances state authorities favor informal
local control instead of public law enforcement. Unravelling these issues is
crucial if scholars wish to better understand how and why states and civil
society decide to use violence.

>>>>

% Packages I've been using for different reasons
\usepackage{hyperref}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usepackage{mathtools}
\usepackage{caption}

@@

\section*{Introduction}
\label{sec:introduction}

@@

The regulation of violence is a crucial function of the state. When states
control violence effectively, markets, politics, and civil society can flourish
\citep{besley2011pillars, north2009violence}. However, states still struggle to
maintain the monopoly of force in much of the Global South. Scholars have
written extensively about how rebel organizations build support to challenge
government institutions, yet we know little about what motivates local
extrajudicial violence and vigilantism in the developing world
\citep{bateson2020politics}.\footnote{On vigilantism more generally, see
\citet{schuberth2013challenging,
zizumbo2017community,smith2019contradictions,cohen2022collective}.} This
omission is significant not only because local violence is pervasive in weak
states, but also because it often comes at a greater social cost than terrorism
or civil wars \citep{blair2017predicting}.

One of the most serious forms of extrajudicial violence is lynching. Lynching
can be defined as ``incidents of physical violence committed by large numbers
of private citizens against one or more individuals accused of having committed
a ``criminal'' offense, whether or not this violence resulted in the death of
the victim(s)'' \citep[645]{godoy2004justice}. Although lynchings occur in more
than one hundred countries in all regions of the world
\citep{jung2020lynching,smith2019contradictions}, Latin America has been
particularly affected by a sharp increase in vigilante violence. Lynching
episodes have been reported in Guatemala, Mexico, Peru, Venezuela and other
countries in the region \citep{barbara2015vigilantes, cruz2019determinants,
godoy2004justice}. From 2011 to 2015, Brazil registered about 2,500 lynching
episodes, and 173 people were killed by angry mobs in 2015 alone---nearly one
execution every two days \citep{barbara2015vigilantes, oliveira2016mob}.
According to José de Souza Martins \citeyearpar{martins2015linchamentos}, who
has studied lynchings in Brazil for more than thirty years, these figures are
not only the highest in the country's history, but among the highest in the
world. The people who participate in lynchings are typically young men, but
they also sometimes include teenage women and girls, elderly women, and even
members of the local police \citep{moura2017linchamentos}.

@@

Our empirical approach offers several distinct advantages. First, the data we
collected come from the same level of analysis as the phenomenon itself: local
residents. Second, online surveys are less susceptible to social desirability
bias, and survey experiments help to elicit sensitive information---an
especially important benefit in researching this topic
\citep{horiuchi2022does}. Third, the experimental nature of the approach allows
us to precisely estimate effect sizes and compare different theoretical
explanations simultaneously. Finally, we can also control for confounding,
which is a potential concern in previous research on vigilantism.

@@

\section*{When Is Lynching Perceived as More Justified?}
\label{sec:exp01}

@@

Figure \ref{fig:exp01} shows the main results for the conjoint experiment. The
graph illustrates the preference associated with each attribute of the
hypothetical lynching episode. Dots with horizontal bars represent point
estimates and 95 percent confidence intervals from linear regressions with
robust standard errors clustered at the respondent level.\footnote{Based on a
semisupervized Dirichlet allocation model, we estimated about 20 percent of
respondents were opposed to lynchings in general but forced to choose. Please
refer to Section C.5 of the Supplementary Material for further information.}

@@

%%%%%%%%%%%%%%%%%%%%%
\section*{How to Reduce Support for Lynchings}
\label{sec:exp03}

Even if people view lynchings as a legitimate activity, that does not mean that
this norm cannot be changed. In our second experiment, we measured the effect
of information provision on attitudes about lynching. More specifically, we
tested whether reminding respondents about the legal and social consequences of
vigilante justice reduces subjects' support for it.\footnote{To prevent
carryover effects caused by previous exposure to the conjoint, we randomized
the order of experiments in our survey.} The experiment included three
treatment conditions and a control group. In all of them we presented
respondents with a short statement affirming that some Brazilians support
vigilantism under certain conditions. We asked the control group to rate their
agreement with the statement. Respondents were asked to use 0 to 49 if they
disagreed, 50 if they neither agreed nor disagreed, and 50--100 if they agreed
with the statement.

@@

\section*{Conclusion}
\label{conclusion}

@@

@@

These findings have implications that extend well beyond Brazil. In particular,
our experiments indicate that lynchings have a crucial gendered aspect, which
has also been documented in other contexts. Using data from 18 Latin American
countries, \citet{nivette2016institutional} also finds that respondents are
most likely to support lynchings when the criminal raped a child, and studies
about lynchings in the American South point out that several episodes resulted
from accusations of sexual assault \citep{jacquet2013giles, smaangs2020race}.
We believe that a culture of honor may explain these results. Individuals in
honor societies view crime as an attack on their personal reputation and, in
turn, are more likely to take revenge to defend their status and that of people
perceived as deserving protection, such as women and children
\citep{nisbett2018culture}. A culture of honor may also explain why Brazilians
see lynchings carried out by the family of the victim as more justified, as
well as refraining from using extralegal violence if it can trigger vendettas.
We expect similar results in societies which share those cultural norms.
Lastly, the fact that race does not appear to be a major motivation behind
lynchings also reflects the experience of places like Haiti or Southern Africa,
where popular violence was mainly driven by other social factors
\citep{berg2011globalizing, jung2020lynching}. Even in the American South, it
Whites and Blacks also lynched people of their own race \citep{beck1997race}.
In this respect, our paper also highlights that lynchings in the Global South
may be more strongly connected with ideas of ``popular justice'' than racial
animus \citep{martins2015linchamentos}.

Our work also suggests new avenues for further research. While previous studies
indicate that there is a relationship between pro-lynching attitudes and actual
engagement in vigilante violence \citep{weisburd1988vigilantism}, future
research should try to identify which factors motivate citizens to participate
in lynchings episodes \citep{nivette2016institutional}. Moreover, it is still
unclear whether other forms of vigilantism, such as security patrols, property
damages, or coercive interrogations have similar levels of popular support
\citep{bateson2020politics}. Finally, lynchings have often been described as a
response to low state capacity \citep{trevizo2022mexico}, yet in some cases
state agents actively incite or engage in vigilante violence themselves
\citep{arias2010violent}. Future studies may investigate in which circumstances
state authorities favor informal local control instead of public law
enforcement. Unravelling these issues is crucial if scholars wish to better
understand how and why states and civil society decide to use violence.
